---
title: "IBD crude incidence calculation"
author: "Mohmad A. Mohamad"
date: "2023-12-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Reference paper**

We tried to simulate the claculation of crude incidence rate for IBD
across different ethnic groups from the following paper:

Misra R, Limdi J, Cooney R, Sakuma S, Brookes M, Fogden E, Pattni S,
Sharma N, Iqbal T, Munkholm P, Burisch J, Arebi N. Ethnic differences in
inflammatory bowel disease: Results from the United Kingdom inception
cohort epidemiology study. World J Gastroenterol. 2019 Oct
28;25(40):6145-6157. doi: 10.3748/wjg.v25.i40.6145. PMID: 31686769;
PMCID: PMC6824277.

**Statistical methodology in reference paper**

Crude incidence rates were calculated according to the following
equation:

Crude incidence rate = No. of cases/ Population at risk

Confidence intervals were calculated assuming a Poisson distribution:

95% CI = Incidence rate ± 1.96 x (Incidence rate / sqrt(Number of
cases))

For example, for the crude IBD incidence rate:

Incidence rate = 17.0 per 100,000, Number of cases = 339 Plugging this
into the formula: 95% CI = 17.0 ± 1.96 x (17.0 / sqrt(339))

95% CI = 17.0 ± 7.3

Giving a 95% confidence interval of 17.0 (7.3 - 26.7)

**Our Statistical methodology**

We retrieved IBD No. of cases and total population at risk from the
reference population and tried to replicate the calculations using EpiR
package

**1. Estimation of crude incidence rates with CI**

```{r IBD crude}
#| echo: TRUE
#| message: false
# Load required packages
library(epiR)
library(ggplot2)
library(readxl)
library(tidyverse)
library(rstatix)
library(ggpubr)

# Load required data
data <- read_xlsx("IBD.xlsx")

# Make Ethnic groups as factor
data$`Ethnic group` <- factor(data$`Ethnic group`, levels = 
                      c("Total population", "White European",
                        "Indian", "Pakistani", "Others"))

# Create a matrix of cases and population at risk
tmp<- as.matrix(cbind(data$Cases,data$`Population at risk`))

# Calculate crude IR, CI
# add calculated column to the data
data <- cbind(data,epi.conf(tmp, ctype = "inc.rate", method = "exact", N = 1000, design = 1, conf.level = 0.95) * 100000)

print(data)

```

**2. Compare incidence rates using pairwise Chi-Square test**

```{r compare}
#| echo: TRUE
#| message: false
# Add column for control
data <- mutate(data, Control= `Population at risk`-Cases)

# Create contingency table
xtab <- data %>% select(Cases, Control)
dimnames(xtab) <- list(`Ethnic group`= data$`Ethnic group`, IBD= c("Cases", "Control"))

# Pairwise comparisons
prop.test <- pairwise_prop_test(xtab)
print(prop.test)
```

**3. Make bar chart**

```{r bar chart,fig.cap='Crude incidence rates across different ethnic groups'}
#| echo: TRUE
#| message: false
# add a column with Ethnic group labels
prop.test <- prop.test %>% mutate(`Ethnic group`= group1)

# Exclude total population from the comparsion
prop.test.short <- prop.test %>% filter(group1 !=" Total population") %>%
  filter(group2 != "Total population")
                                                                                     
# Make a bar chart
ggplot(data, aes(x = `Ethnic group`, y = est, fill=`Ethnic group`)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), position = position_dodge(0.9), width = 0.25) +
  labs(title = "",
       x = "Ethnic Group",
       y = "Crude incidence rate (per 100,000)") +
  geom_text(aes(label = round(est, digits=2)), nudge_y = 1, nudge_x=0.1, hjust="left", angle=45 ) +
theme_minimal() +

# add p values
  stat_pvalue_manual(prop.test.short, label="p = {scales::pvalue(p.adj)}",
  y.position = 38, step.increase = 0.1 , hide.ns = TRUE)

```

```{r print chart}
#| echo: False
#| message: False
png("Crude chart.png",res=300, width = 5500, height= 3500, pointsize=36 )
# Make a bar chart
ggplot(data, aes(x = `Ethnic group`, y = est, fill=`Ethnic group`)) +
  geom_bar(stat = "identity", position = "dodge", color = "black") +
  geom_errorbar(aes(ymin = lower, ymax = upper), position = position_dodge(0.9), width = 0.25) +
  labs(title = "",
       x = "Ethnic Group",
       y = "Crude incidence rate (per 100,000)") +
  geom_text(aes(label = round(est, digits=2)), nudge_y = 1, nudge_x=0.1, hjust="left", angle=45, size=6 ) +

# add p values
  stat_pvalue_manual(prop.test.short, label="p = {scales::pvalue(p.adj)}",
  y.position = 38, step.increase = 0.1, hide.ns = TRUE,
  size= 6, color= "darkblue") +

# adjust text
theme(
  plot.title = element_text(face = "bold", size = 24),
  axis.title.x = element_text(face = "bold", size = 28),
  axis.title.y = element_text(face = "bold", size = 28),
  axis.text.x = element_text(face = "bold", size = 20),
  axis.text.y = element_text(face = "bold", size = 20),
  legend.title = element_text(face = "bold", size = 30),
  legend.text = element_text(face = "bold", size = 28)
  )
dev.off()
```
